BlockTensorDecompositions.jl: A Unified Constrained Tensor Decomposition Julia Package

### 2025-06-19 paper/paper.qmd
add hessian reference and more detailed low rank section

For a tensor input $A$ of order $N$, the Hessian tensor $\nabla^2 f(A)$ is of order $2N$.

@@

## Tensor rank
- tensor rank
- constrained rank (nonnegative etc.)

@@

@@

Most tensor decomposition algorithms require the rank as input [CITE] since calculating the rank of the tensor can be NP-hard in general [@vavasis_complexity_2010]. For applications where the rank is not known a priori, a common strategy is to attempt a decomposition for a variety of ranks, and select the model with smallest rank that still achieves good fit between the factorization and the original tensor.

@@

>>>>

For a tensor input $A$ of order $N$, the Hessian tensor $\nabla^2 f(A)$ is of order $2N$. This contrasts the Hessian *matrix* constructed by vectorizing the function $f$'s input [@magnus_matrix_2019, Ch. 10].

@@

## Tensor rank {#sec-tensor-rank}
- tensor rank
- constrained rank (nonnegative etc.)

@@

By convention, we say the zero tensor $Y=0$ has rank $0$.

@@

Most tensor decomposition algorithms require the rank as input [CITE] since calculating the rank of the tensor can be NP-hard in general [@vavasis_complexity_2010]. For applications where the rank is not known a priori, a common strategy is to attempt a decomposition for a variety of ranks, and select the model with smallest rank that still achieves good fit between the factorization and the original tensor. See section @sec-estimating-tensor-rank for an implementation of this strategy.

@@

### Estimating Tensor Rank {#sec-estimating-tensor-rank}

In many applications, the rank of the input tensor $Y$ may not be known. In the case of CP and Tucker-1 decompositions, there are known bounds on the rank.

::: {#lem-tensor-rank-bounds}

Given a tensor $Y\in\mathbb{R}^{I_1\times \dots \times I_N}$, we have the following bounds on the Tucker-1 and CP rank [@kolda_TensorDecompositionsApplications_2009, Sec. 3.1].

$$
0 \leq \text{rank}_\text{CP} (Y) \leq \min_n\prod_{m\neq n} I_m = \min(I_2\cdots I_N, I_1 I_3 \cdots I_N, \dots, I_1 \cdots I_{N-1}),
$$

and
$$
0 \leq \text{rank}_\text{Tucker-1} (Y) \leq \min\left(I_1, \prod_{n=2}^N I_n\right) = \min(I_1, I_2\cdots I_N) .
$$

We have a rank of $0$ if and only if $Y=0$ is the zero tensor.
:::

Both of these bounds are natural extensions of the typical matrix rank bounds. For a matrix $Y\in\mathbb{R}^{I\times J}$, we know

$$
0 \leq \text{rank}(Y) \leq \min(I, J).
$$

In the case that we try to factorize a matrix $Y$ at a rank $R \geq \text{rank}(Y)$, we should still be able to achieve a final objective of zero $0=\frac{1}{2} \left\lVert A^*B^* - Y \right\rVert_F^2$, and at any rank $R < \text{rank}(Y)$, the objective is positive and decreasing. This fact extends to CP and Tucker-1 decompositions. So a naive strategy for finding the rank of a tensor $Y$, would be first check if it is rank zero, and then try factorizing $Y$ at incrementally larger ranks these two types of decompositions until the final objective hits zero.

This approach has two problems. With real work data, the input tensor is often a noisy version of a low rank tensor $Y = Y_{\text{clean}} + Z$. Noise is often full rank and would cause the input tensor $Y$ to be the maximum rank [@vershynin_HighDimensionalProbability_2018]. Secondly, even if we had a perfectly low rank tensor $Y$, our algorithm at best converges to the optimal solution in the limit. So any finite stopping point has a positive objective.

A simple fix would be to extend the concept of numerical rank (also called $\epsilon$-rank) to tensors.[^numerical-rank-proofs] This serves as an approximation for the rank where taking $\epsilon \rightarrow 0^+$ would return the usual definition of the rank.

[^numerical-rank-proofs]: See <https://mpf-optimization-laboratory.github.io/opt-blog/posts/epsilon-rank/> for an in-depth discussion on this topic. TODO move some proofs to this document.

::: {#def-epsilon-rank}
The $\epsilon$-rank [@golubRosetakDocumentRank1977] of a matrix $A$ is the smallest rank obtainable by an $\epsilon$-perturbation of the matrix $A+E$:
$$
\mathrm{rank}_{\epsilon}(A)=\min_{\left\lVert E \right\rVert\leq \epsilon}\mathrm{rank}(A+E).
$$ {#eq-epsilon-rank}

Here we use $\left\lVert E \right\rVert=\max_{\lVert v \rVert_2=1}\left\lVert Ev \right\rVert_{2}$ to denote the the operator norm.

This definition is *stable* in the sense that adding a small amount of noise to a matrix does not effect the $\epsilon$-rank. This is unlike the traditional rank where adding noise to a low rank matrix can turn it into a full rank matrix, even for an "$\epsilon$" amount of noise. This is clear from @prp-singular-values which also offers a practical method for computing $\mathrm{rank}_{\epsilon}(A)$.

::: {#prp-singular-values}
# Singular Value Characterization

The $\epsilon$-rank of $A$ is equal to the number of singular values of $A$ strictly bigger than $\epsilon$.
:::

The following new result, @thm-rank-recovery, shows that we can use the $\epsilon$-rank to compute the true rank of a matrix under some additive noise $B$ with the right choice of $\epsilon$.

::: {#thm-rank-recovery}
# Rank recovery

Let $A,B\in\mathbb{R}^{m\times n}$ where $\left\lVert B \right\rVert_{2} < \sigma_r / 2$. Then,
$$
\mathrm{rank}_{\epsilon}(A+B)=\mathrm{rank}(A)=r
$$

for all $\epsilon$ between $\left\lVert B \right\rVert_{2} \leq \epsilon < \sigma_{r}/2.$ Here, we list the singular values of $A$ in non-increasing order: $\sigma_1\geq \dots \geq \sigma_r > \sigma_{r+1}=\dots=\sigma_{\min(m,n)}=0$.
:::

Note there are no assumptions on the noise $B$. It could be fixed, or come from any distribution. In the special case that you know the distribution $B$ comes from, you could use that information to estimate an appropriate choice of $\epsilon$. @cor-gaussian-rank-recovery gives the case where $B=Z$ is standard Gaussian noise.

::: {#cor-gaussian-rank-recovery}
# Rank Recovery with Gaussian Noise

Let $Z\in\mathbb{R}^{m\times n}$ be a Gaussian matrix with standard normal entries $Z_{ij}\sim\mathcal{N}(0,1)$ and $\sigma_{r}$ be the largest nonzero singular value of $A\in\mathbb{R}^{m\times n}$. If $\sigma_{r}>2(\sqrt{ m }+\sqrt{ n })$, then for $(\sqrt{ m }+\sqrt{ n } +t) \leq \epsilon <\sigma_{r}/2$, $$\mathrm{rank}_{\epsilon}(A+Z)=\mathrm{rank}(A)$$ with high probability at least $1-2\exp(-ct^2)$ for some constant $0<c\in\mathbb{R}$.
:::

Extending these results to constrained tensors factorization should be possible, but there presents multiple ways we could approach the problem. In the case of identifying rank for a signal decomposition, we are interested the smallest sum of sources under constraints on the allowed sources. For nonnegative sources, the well studied nonnegative rank [@gillis_nonnegative_2020] is most appropriate:

$$
\mathrm{rank}_+(X)= \min \left\{R \in \mathbb{Z}_{+} \middle| \exists A,B\geq 0 \text{ s.t. } X[i,j] = \sum_{r=1}^R A[i,r]B[r,j]\right\}.
$$ {#eq-nonnegative-rank}

Unfortunately, computing this is NP hard in general [@gillis_nonnegative_2020; @vavasis_complexity_2010; @gillis_geometric_2012]. This means extending the $\epsilon$-rank to something like an $\epsilon$-nonnegative-rank may not be practically computable.

Even in the unconstrained case, we have only kicked the problem down the road and must estimate a suitable $\epsilon$ to extract the rank.

There are a number of proposed solutions to this problem that all follow the general principle of Occam's razor of selecting the model with the smallest rank that still achieves a reasonable fit with the data. For automated rank selection, there are many criteria used to balance simplicity with accuracy. The following is a non exhaustive list of methods used in practice in roughly chronological order:

- Akaike [@akaike_new_1974; @burnham_model_1998] and Bayesian [@schwarz_estimating_1978; @neath_bayesian_2012] Information Criterion
- Numerical rank [@golubRosetakDocumentRank1977] and in combination with polynomial filtering [@ubaru_fast_2016]
- Matrix perturbation theory [@ratsimalahelo_rank_2001]
- Consensus clustering and cophenetic correlation coefficient [@monti_consensus_2003;@brunet_metagenes_2004]
- Point of maximum curvature of the approximation error as a function of the rank [@satopaa_finding_2011]
- Cross validation [@austin_tensor_2014]
- "SCREE" plot and segmented linear regression [@saylor_CharacterizingSedimentSources_2019, Sec. 3.2]
- Minimum description length [@fu_model_2019]
- Concordance [@fogel_rank_2023]
- And hypothesis testing [@cai_rank_2023].

The number of ways to estimate rank---and lack of consensus on when to use each method---is best summarized by Fu et. al. [@fu_model_2019]:

> "Even though the selection of rank is very important, in most studies on [non-negative tensor factorization] the value of R is simply determined by trial and error or specialists' insights, and there does not exist a good way to determine R automatically."

We take an approach similar to Satopaa [@satopaa_finding_2011], Saylor [@saylor_CharacterizingSedimentSources_2019, Sec. 3.2], and Graham [@graham_tracing_2025]. We first factorize the tensor at every possible rank (as given by the bounds in @lem-tensor-rank-bounds), and then look at the function that takes as input a rank $r$, and output the final relative error between the reconstructed decomposition at that rank $X_r^*$ and the input tensor $Y$,

$$
f(r) = \left\lVert X_r^* - Y \right\rVert_F / \left\lVert Y \right\rVert_F.
$$

We select the "knee" or "elbow" of this function by finding the point of maximum curvature $\kappa_f(r)$ with finite differences,

$$
\hat{R} = \arg\,\max_r \kappa_f(r) := \frac{f''(r)}{(1+ (f'(x))^2)^{3/2}}.
$$

Since scaling the function $f$ would influence the curvature, we standardize the function by scaling it to the $[0, 1]^2$ box. The following is the specific implementation of the rank finding algorithm, and method for calculating the standard curvature.

Since the maximum rank could be very large, we also include an option `online_rank_estimation` which stops looking are larger ranks if the standard curvature has noticeably dropped, implying a local maximum.

```julia
function rank_detect_factorize(Y; online_rank_estimation=false, rank=nothing, model=Tucker1, kwargs...)
    if isnothing(rank)
        # Initialize output and final error lists
        all_outputs = []
        final_rel_errors = Float64[]

        # Make sure RelativeError is part of the stats keyword argument
        kwargs = isempty(kwargs) ? Dict{Symbol,Any}() : Dict{Symbol,Any}(kwargs)
        get!(kwargs, :stats) do # If stats is not given, populate stats with RelativeError
            [Iteration, RelativeError, ObjectiveValue, isnonnegative(Y) ? GradientNNCone : GradientNorm]
        end
        if RelativeError ∉ kwargs[:stats] # If stats was given, make sure RelativeError is in the list stats
            kwargs[:stats] = [RelativeError, kwargs[:stats]...] # not using pushfirst! since kwargs[:stats] could be a Tuple
        end
        kwargs[:model] = model # add the model back into kwargs

        for rank in possible_ranks(Y, model)
            @info "Trying rank=$rank..."

            kwargs[:rank] = rank # add the rank into kwargs

            output = factorize(Y; kwargs...) # safe to call factorize (rather than _factorize) since both factorize and rank_detect_factorize have checks to see if the keyword `rank` is provided
            push!(all_outputs, output)
            _, stats, _ = output

            final_rel_error = stats[end, :RelativeError]
            push!(final_rel_errors, final_rel_error)
            @info "Final relative error = $final_rel_error"

            if (online_rank_estimation == true) && length(final_rel_errors) >= 3 # Need at least 3 points to evaluate curvature
                curvatures = standard_curvature(final_rel_errors)
                if curvatures[end] ≈ maximum(curvatures) # want the last curvature to be significantly smaller than the max
                    continue
                else
                    # we must have curvature[end] < maximum(curvature) so we can now return
                    R = argmax(curvatures)
                    @info "Optimal rank found: $R"
                    return ((all_outputs[R])..., final_rel_errors)
                end
            end
        end

        # Return if online_rank_estimation == false, or a clear rank was not found
        R = argmax(standard_curvature(final_rel_errors))
        @info "Optimal rank found: $R"
        return ((all_outputs[R])..., final_rel_errors)
    else
        return factorize(Y; rank, model, kwargs...)
    end
end
```

```julia
"""
Approximate first derivative with finite elements. Assumes y[i] = y(x_i) are samples with unit spaced inputs x_{i+1} - x_i = 1.
"""
function d_dx(y::AbstractVector{<:Real})
    d = similar(y)
    each_i = eachindex(y)

    # centred estimate
    for i in each_i[begin+1:end-1]
        d[i] = (-y[i-1] + y[i+1])/2
    end

    # three point forward/backward estimate
    i = each_i[begin+1]
    d[begin] = (-3*y[i-1] + 4*y[i] - y[i+1])/2

    i = each_i[end-1]
    d[end] = (y[i-1] - 4*y[i] + 3*y[i+1])/2
    return d
end

function d2_dx2(y::AbstractVector{<:Real})
    d = similar(y)

    for i in eachindex(y)[begin+1:end-1]
        d[i] = y[i-1] - 2*y[i] + y[i+1]
    end

    # Assume the same second derivative at the end points
    d[begin] = d[begin+1]
    d[end] = d[end-1]
    return d
end

function standard_curvature(y::AbstractVector{<:Real}; kwargs...)
    # An interval 0:10 has length(0:10) = 11, but measure 10-0 = 10
    # hence the length(y) - 1
    Δx = 1 / (length(y) - 1)
    y_max = maximum(y)
    dy_dx = d_dx(y; kwargs...) / (Δx * y_max)
    dy2_dx2 = d2_dx2(y; kwargs...) / (Δx^2 * y_max)
    return @. dy2_dx2 / (1 + dy_dx^2)^1.5
end
```
